Standard Candle Cosmology With Incomplete Spectroscopic Classification

### 2015-11-03 doc/article.tex
new version of document

\section{DES SN Survey Model}
\label{snIamodel:sec}
\subsection{Model}
\label{model:sec}
This Section presents a candidate model
for supernova surveys that has the fidelity needed for the photometric
supernova analysis of the Dark Energy Survey and the Large Synoptic Survey Telescope.
This model captures the current treatment in current analyses but expanded to include
photometry-only transients.  This may not be the ultimate model used, indeed many models
need to be considered for model testing.

The Probabilistic Graphical Model is shown in Figure~\ref{pgm:fig}. The underlying behavior of the universe is described by the following elements:
\begin{description}
\item [Cosmology] Luminosity distance for the cosmological model being considered,
e.g.\ a $wCDM$ universe parameterized by $\Omega_M$, $\Omega_w$, $w_0$, $w_a$.
\item[SNe~Ia, Non-Ia Populations] Time-dependent luminosity for all transient populations
that may enter
the analysis.  The SALT2 model is a candidate for SNe~Ia, with parameters $\alpha$, $\beta$.  The PSNID treatment of non-Ia's can be made into a model for the other
transients.
\item [Relative Rates] Relative rates of all the transients populations and
subpopulations (e.g. SALT2 $x_1$, $c$) under consideration.  The
model could of been extended to measure absolute rates, but that is unnecessary if
the cosmological parameters are the goal.
\end{description}
Each transient is described by the following elements:
\begin{description}
\item[Host Galaxy] The host galaxy, or more generally for the case of hostless supernovae,
the environment from which the transient arises.  The host is described by its redshift
and other parameters that may be of interest such as mass,
specific star formation rate, metallicity.
\item[Type Subtype] Specification of the the transient: the population to which it belongs
and its internal parameters.  For example, an object could be SN~Ia with specific
values for its SALT2 parameters $c$, $x_1$, $t_0$ and its random dispersion
realization (both absolute magnitude and from the error snake). 
This is dependent on the relative rates and
its host galaxy including redshift.
\item[Distance] Distance modulus or equivalently the luminosity distance of the host galaxy,
which in our model corresponds to the cosmological redshift of the transient.  The value
is fixed based on the cosmology and redshift.
\end{description}
The individual photometric measurements of a transient are described by the following
elements:
\begin{description}
\item[Luminosity]  Luminosity as a function of wavelength
corresponding to the date of observation.   This depends on the population, type,
and subtype.  Given evidence for correlations between host properties and Hubble
residuals, a direction connection between host galaxy and luminosity is made.
\item[Flux] Flux of the transient incident to earth at the date of observation.  It is 
fixed based on the luminosity, redshift, and distance.
\end{description}
The observatory is described by the following element:
\begin{description}
\item[Throughput] The transmission function experienced by the transient during
the observation.  It depends on band, position on the focal plane, and date of observation.
\end{description}
The data (subscripted by ``o'' for observation)  are:
\begin{description}
\item[$\text{Photometry}_o$] Photometric flux measurement of the transient.
The expected measurement
is the incident flux integrated over the throughput.
\item[$\text{Type Subtype}_o$] Spectroscopic
measurements of the intrinsic transient type.  It is dependent on the underlying
type.
\item[$\text{Redshift}_o$] Measurement of the host redshift.  The redshift can be
spectroscopic or photometric.  It is dependent on the host redshift, and
redshifts of any other galaxies that may be confused as the host (not shown explicitly
in the PGM).
\item[$\text{Host Properties}_o$] Measurement of the host properties such
as mass, sSFR, metallicity. It is dependent on the host galaxy, and
any other galaxies that may be confused as the host.
\end{description}

@@

\subsection{Sample Selection}
The transients that enter the analysis satisfy some selection criteria.  Trivially, they
have to be discovered in the first place and often more stringent photometric signal-to-noise
requirements are applied.  Additionally, the probability for a transient to have successful
spectroscopy may be dependent on model parameters.

Sample selection criteria for DES have not yet been finalized
(Argonne for photometry, OzDES/Ryan for spectroscopy are candidates to do this);
presented here is a sketch of the likelihood calculation for a reasonable
photometric selection criterion:
for each transient $\nu$, the criterion is to have $\ge N$ photometric points
above a signal-to-noise threshold given $M_\nu$ available measurements.
Each photometric point is associated with a flux threshold that it either passes
or fails.  Given a total $M=\sum_\nu M_\nu$ data points, there are
$2^M$ possible configurations
of passing or failing theshold, of which only
$\prod_\nu \sum_{k=N}^{M_{\nu}} \binom{M_{\nu}}{k}$  configurations lead to
the inclusion of the transients in the sample.

The likelihood for the analysis
sample is the conditional distribution obtained by applying the selection criteria
on the domain of the likelihood given no selection criteria.  This latter likelihood
can be written
as the sum of the contributions from all possible threshold-passing configurations
\begin{equation}
p(data|model) = \sum_{k=1}^{2^M} p(data, \hat{b}_k| model),
\label{likelihood_u:eqn}
\end{equation}
where $\hat{b}$ is the set containing all possible configurations.
Let $b$ be the subset of $\hat{b}$ 
that satisfies sample selection. Then the selection criteria is $\mathcal{S} = \hat{b}_k
\in b$.
The likelihood of the analysis sample is thus proportional to Eqn.~\ref{likelihood_u:eqn}
with failing configurations removed
\begin{equation}
\mathcal{L}(data|\mathcal{S}, model) \propto
\sum_{k=1}^{2^M} p(data, \hat{b}_k| \hat{b}_k
\in b, model) = \sum_{k=1}^{\prod_\nu \sum_{k=N}^{M_{\nu}} \binom{M_{\nu}}{k}} p(data, b_k| model).
\end{equation}

@@

Models for non-Ia supernovae need to be selected.  The model used in PSNID is a good
place to start.

@@

{\bf The non-Ia model needs to be identified.  Perhaps PSNID.}
\item The theoretical distance modulus is
\begin{equation}
\mu(Host, Cosmology) = \mu_{wCDM}(z; \Omega_M, w)
\end{equation}
is the predicted distance modulus of a flat dark energy cosmology
with parameters $\Omega_M$ and $w$, and host redshift $z$.

@@

\section{Tasks}
Here are a list of initial tasks and ingredients needed to proceed.
\begin{itemize}
\item Determination of flux thresholds for each photometric point given
a $S/N$ threshold
\begin{equation}
threshold(point, S/N).
\end{equation}
If we work in the sky-noise dominated regime, I believe this information is already
calculated by DM.
\item Identification of selection criteria for spectroscopic success, as a function
of data.
\item A rates model for transient types and subtypes
\begin{equation}
\pi(z;r).
\end{equation}
\item Decision on how to model the distribution of $x_1$ and $c$.  For a first try a
single normal distribution seems sufficient but do we want to be more ambitious?
\item A non-Ia transient model.  Much of its usage is in classification, so the PSNID
description is a place to start.
\item Calibration model and uncertainties. Assuming we work with zeropoints
$p(Z)$ is needed.
\item Data! Spectroscopic Type and uncertainties, Spectroscopic redshifts and
uncertainties, galaxy properties and uncertainties, host-galaxy associations
and uncertainties, photometry and uncertainties.
\item Determine the mechanics of doing the analysis.  A Bayesian MCMC could produce posteriors, which sampler that best suits our needs? Is a frequentist
analysis technically unfeasible? If redshifts were free to wander I would think so, but maybe
it is possible in this model.
\end{itemize}

>>>>

\section{DES SN Survey Model}
\label{snIamodel:sec}
\subsection{Model}
\label{model:sec}
This Section presents a candidate model
for supernova surveys that has the fidelity needed for the photometric
supernova analysis of the Dark Energy Survey and the Large Synoptic Survey Telescope.
This model captures the current treatment in current analyses and is expanded to include
photometry-only transients.  This may not be the ultimate model used, indeed many models
need to be considered for model testing.

The Probabilistic Graphical Model is shown in Figure~\ref{pgm:fig}. The underlying behavior of the universe is described by the following elements:
\begin{description}
\item [Cosmology] Luminosity distance for the cosmological model being considered,
e.g.\ a $wCDM$ universe parameterized by $\Omega_M$, $\Omega_w$, $w_0$, $w_a$.
\item[SNe~Ia, Non-Ia Populations] Time-dependent luminosity for all transient populations
that may enter
the analysis.  The SALT2 model is a candidate for SNe~Ia, with parameters $\alpha$, $\beta$.  The PSNID treatment of non-Ia's can be made into a model for the other
transients or Natahsa Karpenka is working on a model of non-Type~II core collapse SNe.
\item [Relative Rates] Relative rates of all the transients populations and
subpopulations (e.g. SALT2 $x_1$, $c$) under consideration.  The
model could of been extended to measure absolute rates, but that is unnecessary if
the cosmological parameters are the goal.
\end{description}
Each transient is described by the following elements:
\begin{description}
\item[Host Galaxy] The host galaxy, or more generally for the case of hostless supernovae,
the environment from which the transient arises.  The host is described by its redshift
and other parameters that may be of interest such as mass,
specific star formation rate, metallicity.
\item[Type Subtype] Specification of the the transient: the population to which it belongs
and its internal parameters.  For example, an object could be SN~Ia with specific
values for its SALT2 parameters $c$, $x_1$, $t_0$ and its random dispersion
realization (both absolute magnitude and from the error snake). 
This is dependent on the relative rates and
its host galaxy including its redshift.
\item[Distance] Distance modulus or equivalently the luminosity distance of the host galaxy,
which in our model corresponds to the cosmological redshift of the transient.  The value
is fixed based on the cosmology and redshift.
\end{description}
The individual photometric measurements of a transient are described by the following
elements:
\begin{description}
\item[Luminosity]  Luminosity as a function of wavelength
corresponding to the date of observation.   This depends on the population, type,
and subtype.  Given that searching for correlations between  host properties and Hubble
residuals is a good probe of systematics,
a direction connection between host galaxy and luminosity is made.
\item[Flux] Flux of the transient incident to earth at the date of observation.  It is 
fixed based on the luminosity, redshift, and distance.
\end{description}
The observatory is described by the following element:
\begin{description}
\item[Throughput] The transmission function experienced by the transient during
the observation.  It depends on band, position on the focal plane, and date of observation.
\end{description}
The data (subscripted by ``o'' for observation)  are:
\begin{description}
\item[$\text{Photometry}_o$] Photometric flux measurement of the transient.
The expected measurement
is the incident flux integrated over the throughput.
\item[$\text{Type Subtype}_o$] Spectroscopic
measurements of the intrinsic transient type.  It is dependent on the underlying
type.
\item[$\text{Redshift}_o$] Measurement of the host redshift.  The redshift can be
spectroscopic or photometric.  It is dependent on the host redshift, and
redshifts of any other galaxies that may be confused as the host (not shown explicitly
in the PGM).
\item[$\text{Host Properties}_o$] Measurement of the host properties such
as mass, sSFR, metallicity. It is dependent on the host galaxy, and
any other galaxies that may be confused as the host.
\end{description}

@@

\subsection{Sample Selection}
The transients that enter the analysis satisfy some selection criteria.  Trivially, they
have to be discovered in the first place and often more stringent photometric signal-to-noise
requirements are applied.  Additionally, transients that have
successful spectroscopy may be drawn from an underlying population that differs from the
underlying or photometrically-selected samples.

Sample selection criteria for DES have not yet been finalized
(Argonne Eve for photometry, OzDES/Ryan for spectroscopy are candidates to do this);
presented here is a sketch of the likelihood calculation for a reasonable
photometric selection criterion:
for each transient $\nu$, the criterion is to have $\ge N$ photometric points
above a signal-to-noise threshold given $M_\nu$ available measurements.
Each photometric point is associated with a flux threshold that it either passes
or fails.  Given a total $M=\sum_\nu M_\nu$ data points, there are
$2^M$ possible configurations
of the individual points passing or failing theshold, of which only
$\prod_\nu \sum_{k=N}^{M_{\nu}} \binom{M_{\nu}}{k}$  configurations lead to
the inclusion of the transients in the sample.

The likelihood for the analysis
sample is the conditional distribution obtained by applying the selection criteria
on the domain of the likelihood given no selection criteria.  This latter likelihood
can be written
as the sum of the contributions from all possible threshold-passing configurations
\begin{equation}
p(data|model) = \sum_{k=1}^{2^M} p(data, \hat{b}_k| model),
\label{likelihood_u:eqn}
\end{equation}
where $\hat{b}$ is the set containing all possible configurations.
Let $b$ be the subset of $\hat{b}$ 
that satisfies sample selection. Then the selection criteria is $\mathcal{S} = \hat{b}_k
\in b$.
The likelihood of the analysis sample is proportional to Eqn.~\ref{likelihood_u:eqn}
with failing configurations removed
\begin{equation}
\mathcal{L}(data|\mathcal{S}, model) =
\sum_{k=1}^{2^M} p(data, \hat{b}_k| \hat{b}_k
\in b, model) \propto \sum_{k=1}^{\prod_\nu \sum_{k=N}^{M_{\nu}} \binom{M_{\nu}}{k}} p(data, b_k| model),
\end{equation}

@@

Models for non-Ia supernovae need to be selected.

@@

{\bf The non-Ia model needs to be identified.}
\item The theoretical distance modulus is
\begin{equation}
\mu(Host, Cosmology) = \mu_{wCDM}(z; \Omega_M, w)
\end{equation}
is the predicted distance modulus of a flat dark energy cosmology
with parameters $\Omega_M$ and $w$, and host redshift $z$.

@@

\section{Tasks}
Here are a list of initial tasks and ingredients needed to proceed.
\subsection{Photometric Selection}
Here are desired properties of the set (or sets) of objects obtained from photometric selection.
\begin{itemize}
\item Sufficient data-quality to have a chance at photometric classification.
SNe~Ia do look different from other transients: the principal source of type  ambiguity
comes from poor signal-to-noise and poor temporal/wavelength sampling.
\item A large set of likely SNe~Ia.
\item All objects with photometry-only must have spectroscopically-typed analogs, preferably though not necessarily at the same redshift.  SNe~Ia subpopulations whose characteristics
are unconstrained are not useful for cosmology.
\item Obvious non-Ia's from populations that could be misconstrued as SN~Ia.  For example,
a non-Ia at low redshift could be
mistaken to be SN~Ia at high redshift due to degraded signal.
\item Exclusion of objects that are not part of populations that could be misconstrued as SN~Ia.
They would pollute inference of the non-Ia population.
\end{itemize}

A clean deliverable would be:
\begin{itemize}
\item Determination of flux thresholds for each photometric point given
a $S/N$ threshold
\begin{equation}
threshold(point, S/N).
\end{equation}
If we work in the sky-noise dominated regime, I believe this information is already
calculated by DM.
\end{itemize}
\subsection{Others}

\begin{itemize}
\item Identification of selection criteria for spectroscopic success, as a function
of data.
\item A rates model for transient types and subtypes
\begin{equation}
\pi(z;r).
\end{equation}
\item Decision on how to model the distribution of $x_1$ and $c$.  For a first try a
single normal distribution seems sufficient but do we want to be more ambitious?
\item A non-Ia transient model.  Much of its usage is in classification, PSNID or
Karpenka could be places to start.
\item Calibration model and uncertainties. Assuming we work with zeropoints
$p(Z)$ is needed.
\item Data! Spectroscopic Type and uncertainties, Spectroscopic redshifts and
uncertainties, galaxy properties and uncertainties, host-galaxy associations
and uncertainties, photometry and uncertainties.
\item Determine the mechanics of doing the analysis.  A Bayesian MCMC could produce posteriors, which sampler that best suits our needs? Is a frequentist
analysis technically unfeasible? If redshifts were free to wander I would think so, but maybe
it is possible in this model.
\end{itemize}
